Comparing Early Evolution of Flames in X-ray Bursts in Two and Three Dimensions

### 2023-05-17 xrb3/paper.tex
some updates for the refree

\documentclass[linenumbers]{aastex631}

@@

In \citet{eiden:2020}, we developed the simulation methodology in
\castro\ for modeling laterally spreading flames through the accreted
layer on the surface of a neutron star.  For these models,
\castro\ evolves the compressible Euler equations with a Coriolis
force arising from rotation, constant gravity (the plane-parallel
approximation), reactive sources, and thermal diffusion (integrated
explicitly).  We use a general stellar equation of state with
radiation, ions as a perfect gas, and degenerate/relativistic
electrons described in \citet{timmes_swesty:2000}.  Building on these
first simulations, in \citet{harpole:2021}, we explored several
different rotation rates and initial model thermal structures to
understand how the flame speed depends on the thermal structure of the
envelope.  Both of these papers used a two-dimensional axisymmetric
geometry.  The three-dimensional simulations we present here build on
these works.  We use a neutron star crust temperature of $2\times
10^8~\mathrm{K}$ and rotation rate of 1000~Hz, as followed in
\citet{harpole:2021}.  This model was chosen here since it gives a
clean well-defined flame.  The model is constructed following the same
procedure described in \citet{eiden:2020,harpole:2021}.

@@

\castro\ uses the the \amrex\ adaptive mesh refinement library
\citep{amrex_joss} to manage the discretization and domain
decomposition.  Our 3D simulation used a base grid of $768^2 \times
192$ zones with 2 levels of refinement, the first jumping by a factor
of 4 and the next by a factor of 2.  We subcycle in time, so the
coarser grids take a large timestep than the fine grids.  We used
static mesh refinement, to achieve better load balancing, fully
refining the atmosphere below a height of $3600~\mathrm{cm}$.  The
domain has a size $(1.96608\times 10^5~\mathrm{cm})^2 \times
2.4576\times 10^4~\mathrm{cm}$.  The simulation methodology and
parameters are mostly identical to those used in \citet{harpole:2021}.
We use an unsplit piecewise parabolic method
\citep{ppmunsplit,millercolella:2002} for the advection and operator
(Strang) splitting for the reactions, evolving an internal energy
evolution equation during the burn, as described in
\citet{strang_rnaas}.  These simulations used the 7-isotope He-burning
network described in \citet{iso7}.  This was one of the networks used
in \citet{eiden:2020}.  The main difference with the previous
simulations is that we switched from a hydrostatic to reflecting
boundary at the bottom of the domain and used a simple well-balanced
scheme (similar to \citealt{kappeli:2016} but adapted to deal with
characteristic tracing in PPM as described in \citealt{ppm-hse}).
Tests demonstrated that this boundary does a better job than our
previous approach at supporting the atmosphere as the flame propagates
for long times.

@@

\begin{figure}[t]
\centering
\plotone{full}
\caption{\label{fig:vr_abar} Volume rendering of the three-dimensional XRB
simulation showing the mean molecular weight, $\bar{A}$.  Two views
are shown: the left column is viewed from above while the right column
shows it from a shallow angle above the surface.  The panels show the state at 10 ms, 20 ms, and 40 ms (from top to bottom).
The colorbar shows the value of $\bar{A}$ together with the opacity assigned for
that value---this allows us to bring out some structure in the rendering.  A
small triad in the upper right of the top panels shows the orientation,
with red = $x$, green = $y$, and blue = $z$, our vertical direction.}
\end{figure}

Figure~\ref{fig:vr_abar} shows the a three-dimensional simulation via
top-down volume rendering of the mean-molecular weight,
\begin{equation}
\frac{1}{\bar{A}} = \sum_k \frac{X_k}{A_k}
\end{equation}
where $X_k$ is the mass fraction of species $k$ and $A_k$ is the
atomic weight (in mass units).  The structure is shown at 10, 20, and
40~ms of simulation time.  While initially quite symmetric, the axial
symmetry is broken and the flame structure takes on a more tenuous
form.  The panels show the flame spreading considerably through the
domain as time evolves, and the simulation is halted once the flame
nears the boundaries.
\begin{figure*}[t]
\centering
\plotone{time_series}
\caption{\label{fig:2d_abar} Time-sequence of the two-dimensional axisymmetric
simulation using the same setup as the three-dimensional simulation shown
in Figure~\ref{fig:vr_abar}.  The domain is zoomed in, showing only the left
two-thirds of the surface.}
\end{figure*}

\begin{figure*}[t]
\centering
%\epsscale{0.75}
\plotone{flame_wave_1000Hz_25cm_smallplt174562_slice}
%\epsscale{1.0}
\caption{\label{fig:3dslice} A vertical slice through the three-dimensional
simulation at 40~ms showing a region centered on the flame viewed from the side.  The
flame structure is clearly seen. }
\end{figure*}

>>>>

\documentclass[linenumbers,trackchanges]{aastex631}

@@

In \citet{eiden:2020}, we developed the simulation methodology in
\castro\ for modeling laterally spreading flames through the accreted
layer on the surface of a neutron star.  For these models,
\castro\ evolves the compressible Euler equations with a Coriolis
force arising from rotation, constant gravity (the plane-parallel
approximation), reactive sources, and thermal diffusion (integrated
explicitly)\edit1{:
\begin{eqnarray}
\frac{\partial \rho}{\partial t} + \nabla \cdot (\rho \Ub) &=& 0 \\
\frac{\partial (\rho X_k)}{\partial t} + \nabla \cdot (\rho \Ub X_k) &=& \rho \omegadot_k \\
\frac{\partial (\rho \Ub)}{\partial t} + \nabla \cdot (\rho \Ub \Ub) + \nabla p &=& \rho {\bf g} - 2 \rho {\bf \Omega} \times \Ub \\
%% no centrifugal- \rho {\bf \Omega} \times ({\bf \Omega} \times {\bf r}) \\
\frac{\partial (\rho E)}{\partial t} + \nabla \cdot (\rho \Ub E + p \Ub) &=& \rho \Ub \cdot {\bf g} + \nabla \cdot k_\mathrm{th} \nabla T + \rho \dot{\epsilon}
%% no centrifugal - \rho (\Omegab \cdot \rb)(\Omegab \cdot \Ub) + \rho | \Omegab |^2 (\Ub \cdot r)
\end{eqnarray}
Here, $\rho$ is the mass density, $X_k$ are the mass fractions of the
nuclear species with associated creation rates $\omegadot_k$, $\Ub$
the velocity, $E$ is the specific total energy, $p$ is the pressure,
${\bf g}$ is the gravitational
acceleration, $\Omegab$ is the rotation vector, $k_\mathrm{th}$ is the thermal conductivity, and $T$ is the temperature.
}
 We use a general stellar equation of state with
radiation, ions as a perfect gas, and degenerate/relativistic
electrons described in \citet{timmes_swesty:2000} \edit1{which closes
  the system:}
\begin{eqnarray}
E &=& e + \frac{1}{2} |\Ub|^2 \\
p &=& p(\rho, e, X_k) \\
T &=& T(\rho, e, X_k)
\end{eqnarray}
Building on these
first simulations, in \citet{harpole:2021}, we explored several
different rotation rates and initial model thermal structures to
understand how the flame speed depends on the thermal structure of the
envelope.  Both of these papers used a two-dimensional axisymmetric
geometry.  The three-dimensional simulations we present here build on
these works.  We use a neutron star crust temperature of $2\times
10^8~\mathrm{K}$ and rotation rate of 1000~Hz, as followed in
\citet{harpole:2021}.  This model was chosen here since it gives a
clean well-defined flame.  The model is constructed following the same
procedure described in \citet{eiden:2020,harpole:2021}.

@@

\castro\ uses the the \amrex\ adaptive mesh refinement library
\citep{amrex_joss} to manage the discretization and domain
decomposition.  Our 3D simulation used a base grid of $768^2 \times
192$ zones with 2 levels of refinement, the first jumping by a factor
of 4 and the next by a factor of 2.  We subcycle in time, so the
coarser grids take a large timestep than the fine grids.  We used
static mesh refinement, to achieve better load balancing, fully
refining the atmosphere below a height of $3600~\mathrm{cm}$.  The
domain has a size $(1.96608\times 10^5~\mathrm{cm})^2 \times
2.4576\times 10^4~\mathrm{cm}$.  The simulation methodology and
parameters are mostly identical to those used in \citet{harpole:2021}.
We use an unsplit piecewise parabolic method
\citep{ppmunsplit,millercolella:2002} for the advection and operator
(Strang) splitting for the reactions, evolving an internal energy
evolution equation during the burn, as described in
\citet{strang_rnaas}.  These simulations used the 7-isotope He-burning
network described in \citet{iso7}.  This was one of the networks used
in \citet{eiden:2020}.  \edit1{Thermal diffusion is treated explicitly, using the conductivities from \citet{Timmes00}.
The overall integration strategy is second-order accurate in space and time.}  The main difference with the previous
simulations is that we switched from a hydrostatic to reflecting
boundary at the bottom of the domain and used a simple well-balanced
scheme (similar to \citealt{kappeli:2016} but adapted to deal with
characteristic tracing in PPM as described in \citealt{ppm-hse}).
Tests demonstrated that this boundary does a better job than our
previous approach at supporting the atmosphere as the flame propagates
for long times.

@@

\begin{figure}[t]
\centering
\plotone{full}
\caption{\label{fig:vr_abar} Volume rendering of the three-dimensional XRB
simulation showing the mean molecular weight, $\bar{A}$.  \edit1{Two views
are shown: the left column is viewed from above while the right column
shows it from a shallow angle above the surface.  The panels show the state at 10 ms, 20 ms, and 40 ms (from top to bottom).
The colorbar shows the value of $\bar{A}$ together with the opacity assigned for
that value---this allows us to bring out some structure in the rendering.  A
small triad in the upper right of the top panels shows the orientation,
with red = $x$, green = $y$, and blue = $z$, our vertical direction.}}
\end{figure}

Figure~\ref{fig:vr_abar} shows the a three-dimensional simulation via
top-down volume rendering of the mean-molecular weight,
\begin{equation}
\frac{1}{\bar{A}} = \sum_k \frac{X_k}{A_k}
\end{equation}
where $X_k$ is the mass fraction of species $k$ and $A_k$ is the
atomic weight (in mass units).  The structure is shown at 10, 20, and
40~ms of simulation time.  While initially quite symmetric, the axial
symmetry is broken\edit1{, likely due to roundoff error,}  and the flame structure takes on a more tenuous
form.  The panels show the flame spreading considerably through the
domain as time evolves, and the simulation is halted once the flame
nears the boundaries.
\begin{figure*}[t]
\centering
\plotone{time_series}
\caption{\label{fig:2d_abar} Time-sequence of the two-dimensional axisymmetric
simulation using the same setup as the three-dimensional simulation shown
in Figure~\ref{fig:vr_abar}.  The domain is zoomed in, showing only the left
two-thirds of the surface.}
\end{figure*}

\begin{figure*}[t]
\centering
%\epsscale{0.75}
\plotone{time_series_3D}
%\epsscale{1.0}
\caption{\label{fig:3dslice} Vertical slices through the three-dimensional
simulation \edit1{at the same times as in Figure~\ref{fig:2d_abar}}, showing a region centered on the flame viewed from the side.  The
flame structure is clearly seen. }
\end{figure*}
